A Mechanised Library for Generic (Co)Programming

### 2024-03-05 paper/paper.tex
Coalgebra completed.

For example, for a type of states \coq{X} and an alphabet type \coq{L} we can
define a (labelled) transition system (LTS) on \coq{X} as a
function \coq{c : X -> L * X} implementing the \emph{transition} map. In
particular, for a state \coq{x1 : X}, \coq{c x1} returns a
pair \coq{(l, x2) : L * X} where \coq{l : L} is the observable action
and \coq{x2 : X} is the next state. We could then add the following notation
\mpav{is this notation duable in Cock?}
\[
  x_{1} \xrightarrow{l} x_{2} \eqiff c x_{1} = (l, x_{2})
\]
\begin{coqcode}
Notation x1 --l> x2 := c(x1) = x2.
\end{coqcode}

Terminal coalgebras are defined analogously to initial algebras (see
Figure~\ref{fig:term_coalg}).

\begin{coqcode}
CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
\end{coqcode}

We define smart constructors for the isomorphism of greatest fixed points
\begin{coqcode}
g_in : App F GFix ~> GFix
g_out : GFix ~> App F GFix
\end{coqcode}
% \caption{Terminal Coalgebras}\label{fig:term_coalg}
% \end{figure}

Their equivalence relation, and anamorphisms, and universal properties are
easy to encode, in an analogous way to catamorphisms and initial algebras:
\begin{coqcode}
  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...

  Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
This definition, however, prevents us from composing catamorphisms and
anamorphisms: we cannot mix \coq{GFix} and \coq{LFix}.
To address this issue, we consider another alternative definition
for initial $F$-algebras.

>>>>

The seed function can be modelled by a coalgebra. For example, for a type of
states \coq{X} and an alphabet type \coq{L} we can define a (labelled)
transition system (LTS) on \coq{X} as a function \coq{c : X -> L * X}
implementing the \emph{transition} map. In particular, for a
state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.
Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
carrier \coq{X} is a map \coq{X -> FX}. In the previous example \coq{c} is
an \coq{F}-coalgebra on the carrier \coq{X} with \coq{FX = L * X}.

Now an anamorphism takes such a seed function and runs it while collecting all
the observable behaviours. In order to do so we define the type of infinite
streams of labels which is the greatest fixed-point for the functor \coq{A * X}.
We define this in Coq as follows:
\begin{coqcode}
CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
\end{coqcode}
where \coq{GFix_in}  and \coq{GFix_out} are the two inverses witnessing the
isomorphism. Additionally, we define smart constructors for the isomorphism of greatest fixed points
\begin{coqcode}
g_in : App F GFix ~> GFix
g_out : GFix ~> App F GFix
\end{coqcode}

What witnesses the fact that this is the greatest fixed-point is the universal
property of the anamorphism. For a given \coq{F}-coalgebra this is the unique
map that runs the seed function and composes the information from the first step
the result

\begin{coqcode}
Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
: f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
Equivalently, we could have define an anamorphism as the
unique \coq{F}-coalgebra homomorphism, that is a map of the following type
\begin{coqcode}
Coalg F A ~> A ~> GFix
\end{coqcode}
